---
title: "11: Analysis of correlated data"
subtitle: "R 4 ASME"
authors: Andrea Mazzella [(GitHub)](https://github.com/andreamazzella)
output: html_notebook
---

-------------------------------------------------------------------------------

## Contents
* Analysis of correlated data with Poisson and logistic regression
 - robust standard errors
 - Generalised Estimating Equations
 - Random Effects models

-------------------------------------------------------------------------------

## 0. Packages and options

```{r}
# install.packages("miceadds")    # robust standard errors
# install.packages("lme4")        # RE
# install.packages("broom")       # improve regression output
# install.packages("broom.mixed") # improve regression output for lme4
# install.packages("geepack")     # GEE
```


```{r}
# Load packages
library("haven")
library("magrittr")
library("summarytools")
library("survival")
library("miceadds")    # robust standard errors
library("lme4")        # RE
library("broom")       # improve regression output
library("broom.mixed") # improve regression output for lme4
library("geepack")     # GEE
library("tidyverse")

# Limit significant digits to 3, remove scientific notation
options(digits = 3, scipen = 9)
```

-------------------------------------------------------------------------------


# Part 1: Poisson regression

These data are from a pneumococcal vaccine trial performed in Papua New Guinea, assessing the vaccine efficacy in preventing clinical episodes of pneumonia among children.

Each child might have more than one record, because each record represents an episode of pneumonia (or the last period of follow-up, without pneumonia). (This means that the dataset was produced with Stata's `stset` command).

```{r}
papua <- read_dta("pngnew.dta")
glimpse(papua)
```

Variables that we will use:

**Outcome*: `any` (episode of clinical pneumonia during this period)
**Exposure*: `vacc` (vaccination: 1 = placebo, 2 = vaccine)
**Cluster*: `id` (child)
**Time*:
 -`timein` (date of entry in this follow-up period)
 -`timeout` (date of exit from this follow-up period)
 -`dob` (date of birth)
**Other*
- `sex` (1 = male, 2 = female)
- `anyprev` (0: no previous episodes of pneumonia, 1: any prev. episodes)

Label values. The Stata practical will ask you to calculate person-years; let's do it now.
```{r data_management}
papua %<>% mutate(sex = factor(sex, levels = c(1, 2), labels = c("male", "female")),
                 vacc = factor(vacc, levels = c(1, 2), labels = c("placebo", "vaccine")),
                 pyrs = as.numeric(timeout - timein) / 365.25) %>% 
           select(-"datevac")

summary(papua)
```

## 1. Explore the data format

Identify all records for child with ID 2921 and make sure you understand what each row represents.
```{r}
papua %>% filter(id == 2921)
```


## 2. Explore the numbers

```{r}
pap_summ <- papua %>% group_by(id) %>% summarise(episodes = sum(any),
                                                 vacc = max(as.numeric(vacc)))

pap_summ %$% ctable(episodes, vacc, headings = F, prop = "no")
```
A total of 1390 children, 671 of whom are vaccinated. 467 of them did not have any episodes of pneumonia. Two children had 11 episodes of pneumonia, and they were both unvaccinated.


Count the total number of episodes in each intervention arm.
```{r}
papua %>% group_by(vacc) %>% summarise(episodes = sum(any))
```


## 3. Prepare for cohort analysis

Create a survival object to do person-time calculations. Unlike `stset`, `Surv()` doesn't seem to require any special option to include repeated observations.

```{r}
#surv_papua <- papua %$% Surv(time = as.numeric(timein) / 365.25,
 #                            time2 = as.numeric(timeout) / 365.25,
  #                           event = any)

surv_papua <- papua %$% Surv(time = pyrs, event = any)

summary(surv_papua)
```


## 4. Incidence rates and HR (invalid)

Calculate incidence rates in the vaccine and placebo arms, and calculate a rate ratio (without accounting for within-child clustering).

```{r}
pyears(surv_papua ~ vacc, data = papua, scale = 1) %>%
  summary(n = F, rate = T, ci.r = T, scale = 100)

print("Rate ratio")
89.5/99.0
```
*Issue* I don't know how to calculate the HR automatically, or how to get a 95% CI, or a p-value.

From this (incorrect) analysis, the vaccine has a (mild) effect. (NB: from `stmh`: p = 0.02)


## 5a. Poisson ignoring clustering (invalid)

Let's demonstrate that ordinary Poisson regression, ignoring clustering, is also invalid.

Fit an ordinary Poisson regression model for the effect of vaccination.

`broom::tidy()` can be used to improve the output of a `glm()` – it's an alternative to `epiDisplay::idr.display()`.

```{r}
# Poisson model
pois_inv <- glm(any ~ vacc + offset(log(pyrs)),
                family = "poisson",
                data = papua)
# HR and 95% CI
tidy(pois_inv, exponentiate = T, conf.int = T)
?tidy.glm

# log(SE)
summary(pois_inv)
```
HR 0.9 (0.83, 0.98), Wald's p = 0.02
HR is the same as the one calculated with `pyears()`.
SE of log(HR) = 0.04235


## 5b. Poisson with robust standard errors

Now fit again the Poisson model, but take clustering into account by computing robust standard errors. How does this change the estimations of the vaccine efficacy?

We can use function `glm.cluster()` from package {miceadds}. It's similar to `glm()`, and it includes a cluster option (careful – variable needs to be put in quotes)
```{r}
pois_rob <- glm.cluster(papua,
                        any ~ vacc + offset(log(pyrs)),
                        cluster = "id",
                        family = "poisson")

# HR
coef(pois_rob) %>% exp()

# 95% CI
confint(pois_rob) %>% exp()

# SE of log(HR)
summary(pois_rob)
```
HR 0.90: same as above
95% CI: 0.80-1.02 (wider)
SE of log(HR): 0.061
p = 0.10: only weak evidence for an association now.


## 6. Poisson with random effects

Now use a random effects model to account for within-child clustering.

We use `lme4::glmer()`. The overall syntax is the same as `glm()`; you specify the clustering factor like this: `+ (1|clustering_factor)`. (Statistically, this is a "scalar random effect term": it generates 1 random effect per cluster).

The output can be much simplified with `broom.mixed::tidy()`.

```{r}
# Fit model
pois_re <- glmer(any ~ vacc + offset(log(pyrs)) + (1|id),
                 data = papua,
                 family = "poisson")
# Output
tidy(pois_re,
     conf.int = TRUE,
     exponentiate = TRUE,
     effects = "fixed")
```
HR: 0.89 (0.78-1.01) (p = 0.06)
SE: 0.057

*Minor issue* Stata also provides a measure of clustering, theta, and a LRT on the evidence of clustering. Not sure how to extract the theta from the `glmer()` output.
(Stata - theta: 0.76, LRT: p < 0.001)

## 7. Vaccine efficacy

In Stata, this is `xtpoisson` instead of `streg, shared(cluster)`. `xtpoisson` estimates a parameter α, whilst `streg` estimates a parameter θ (but their values are the same).

What is the estimated vaccine efficacy, with an appropriate 95% CI?

```{r}
# Values from regression model
rate_ratio <- 0.886
rr_conf_low <- 0.780
rr_conf_high <- 1.006

# Calculate VE
(vaccine_efficacy <- (1 - rate_ratio) * 100)
(ve_conf_low <- (1 - rr_conf_high) * 100)
(ve_conf_high <- (1 - rr_conf_low) * 100)
```


## 8. Age time-scale

Now set the time-scale as "age", and refit the Poisson model with random effects.

```{r}
surv_papua_age <- papua %$% Surv(time = as.numeric(timein) / 365.25,
                                 time2 = as.numeric(timeout) / 365.25,
                                 event = any,
                                 origin = as.numeric(dob) / 365.25)
```

*-----*
*ISSUE* How do I use the new surv object in the glmer() function? 
*-----* Do I need to offset for something other than pyrs?

```{r}
# Fit model
pois_re_age <- glmer(any ~ vacc + offset(log(pyrs)) + (1|id),
                 data = papua,
                 family = "poisson")
# Output
tidy(pois_re_age,
     conf.int = TRUE,
     exponentiate = TRUE,
     effects = "fixed")
```
>Stata
```{stata}
streg i.vacc, dist(exp) frailty(gamma) shared(id) forceshared
```


## 9. Random effects Poisson model with covariates

Using the survival object with follow-up timescale, fit a random effects model to assess vaccine efficacy controlling for age at the start of each period, and sex.

Let's first calculate age at start, and divide it into categories.

```{r}
# Calculate age at start
papua %<>% mutate(age_start = as.numeric(timein - dob)/365.25)

# Summarise age at start
summary(papua$age_start)
ggplot(papua, aes(age_start)) + geom_histogram()
```
The age at start range from 0.21 to 7.44 years. In the Stata practical, they decide to categorise this in 1-year groups, except the children aged 4 years or more, who are grouped togther.

```{r}
# Categorise age at start
papua %<>% mutate(agegrp = cut(age_start,
                              breaks = c(0, 1, 2, 3, 4,+Inf),
                              labels = c("0-1yr,", "1-2yr", "2-3yr", "3-4yr", ">=4yr")))

# Check it worked
papua %>% group_by(agegrp) %>% summarise(count = n(),
                                         min_age = min(age_start),
                                         max_age = max(age_start))

```

We can now fit the RE model with sex and age as (categorical) covariates.

Does this change the estimates compared with Q6? If so, why?

```{r}
# Fit model
pois_re_age <- glmer(any ~ vacc + agegrp + sex + offset(log(pyrs)) + (1|id),
                 data = papua,
                 family = "poisson")
# Output
tidy(pois_re_age,
     conf.int = TRUE,
     exponentiate = TRUE,
     effects = "fixed")
```
Effect of vaccine vs placebo:
Adjusted:   HR 0.91 (0.82-1.02), p = 0.10
Crude (Q6): HR 0.89 (0.78-1.01), p = 0.06

Controlling for these baseline factors hasn't changed much the estimates – which makes sense, because this is a randomised trial. You can however still check for baseline differences if you want:
```{r}
papua %>% filter(anyprev == 0) %$% ctable(sex, vacc, prop = "c")
papua %>% filter(anyprev == 1) %$% ctable(agegrp, vacc, prop = "c")
```

-------------------------------------------------------------------------------


# Part 2: Logistic regression

These data are from a study on household contacts of tubercolosis cases. We will assess the effect of duration of cough in the index case on the odds of positive Mantoux test in the contacts. Since there is no time element, we will use logistic regression.

## 10. Import and explore data

```{r}
tb <- read_dta("hhtb.dta") %>% mutate_if(is.labelled, as_factor)
glimpse(tb)
```

Variables that we will use:

**Outcome*: `mantoux` (tuberculin test result: 0 = negative, 1 = positive)
**Exposure*: `cough` (duration of cough in index case: 1 = <2 months, 2 = >=2 months)
**Cluster*: `id` (household, so = index case)
**Other*
- `hiv` (HIV status of index case: 1 = negative, 2 = positive)
- `agegrp` (age of contact, in years)

There are value labels in Stata, I'm not sure why haven can't import them. Let's add them, and remove variables we don't need. I'll add an "ix" to variables that relate to the index case because I get confused otherwise.

```{r}
tb %<>% mutate(ix_cough = factor(cough, levels = c(1, 2), labels = c("<2mo", ">=2mo")),
               ix_hiv = factor(hiv, levels = c(1, 2), labels = c("neg", "pos"))) %>% 
  select(-c("smear1", "crowding", "intimacy", "tbsite", "cavit", "cough", "hiv"))

summary(tb)
```

## 11. Explore clusters by HIV status of index case

Summarise the dataset by index case to explore the distribution of number of contacts stratified by HIV status of the index case.

```{r}
# Create summary data
tb_summ <- tb %>% group_by(id) %>%
                  summarise(n_contacts = n(),
                            index_HIV = mean(as.numeric(ix_hiv)))

# Crosstabulate
tb_summ %$% ctable(n_contacts, index_HIV, prop = "n")
```
The dataset contains a total of 70 index cases, 28 of whom were HIV negative and 42 HIV positive.
8 index cases had only 1 contact in their household; 9 had 2, etc.


## 12. Explore clusters by cough of index case

Do the same but stratifying by duration of cough in the index case.

```{r}
# Create summary data
tb_summ <- tb %>% group_by(id) %>%
                  summarise(n_contacts = n(),
                            index_cough = mean(as.numeric(ix_cough)))

# Crosstabulate
tb_summ %$% ctable(n_contacts, index_cough, prop = "n")
```
32 index cases had a cough for 2 months or less, 26 for more than that; for 12, this information had not been recorded.
	 

## 13. Crude analysis ignoring clustering (invalid)

Examine the distribution of positive Mantoux among contacts by the duration of cough in the index case, ignoring any clustering.

What's the OR and 95% CI? χ² p-value? What would you conclude?

```{r}
tb %$% ctable(mantoux, ix_cough, prop = "c", OR = T, chisq = T)
```
Ignosing clustering, 67% of positive people had an index case who had a cough for more than 2 months.
OR 1.78 (1.06-3.01), χ² p = 0.04
With this (invalid) analysis, there is evidence of an association between being a contact of someone with cough of long duration and testing positive for TB.


## 14. Logistic regression with robust standard errors

Fit a logistic regression model that accounts for clustering by calculating robust standard errors.

What conclusions do you derive? Compare this with the above output.

```{r}
# Fit the model
logit_rob <- glm.cluster(tb,
                         mantoux ~ ix_cough,
                         cluster = "id",
                         family = "binomial")
# HR
coef(logit_rob) %>% exp()

# 95% CI
confint(logit_rob) %>% exp()

# SE of log(HR)
summary(logit_rob)
```
OR 1.78 (0.90-3.52), p = 0.10: OR is the same, but CI is wider, and there is less evidence of an association.


## 15. Logistic regression with GEE

Now fit the same logistic model but with Generalised Estimating Equations approach, with robust standard errors and an exchangeable correlation matrix.

Compare the results with the ones above.

We can do this with `geepack::geeglm()`. The correlation matrix type goes into the "corstr" option. The standard errors are automatically calculated as robust.

Note that this is a bit fiddly because the function requires the explanatory variables to not contain any missing data, so you  need to `drop_na()` your dataset.

```{r}
# Fit GEE logistic model
? geeglm
tb_gee <- geeglm(mantoux ~ ix_cough,
                 data = drop_na(tb, ix_cough),
                 id = id,
                 family = "binomial",
                 corstr = "exchangeable")

# Output
tidy(tb_gee,
     conf.int = TRUE,
     exponentiate = TRUE)
```
OR 1.88 (0.95-3.71)
p = 0.07


## 16. Logistic regression with RE and quadrature check

Now do the same but with a random effects model, and do a LRT of rho = 0. Is there evidence of variation between households?

(*Technical note*: there is now a new option, nAGQ = 12. This is because, by default, glmer() uses the Laplace approximation to the log likelihood (1 point per axis), whilst Stata uses the Gauss-Hermite quadratic approximation, with 12 points per axis; this is why you need to use `quadchk` in Stata with this model, so that you can ensure that this quadrature reasonable. Changing nAGQ slightly changes your CI and p-value. For some reason, in Stata the RE model in Q9 does not use quadrature, so even if you try and use `quadchk` it doesn't let you. Is it because that model has covariates? Or because it's a Poisson model? I don't know.)

*Issue*: as earlier, I don't know how to do a LRT of rho = 0.

```{r}
# Fit model
logit_re_age <- glmer(mantoux ~ ix_cough + (1|id),
                 data = tb,
                 family = "binomial",
                 nAGQ = 12)
# Output
tidy(logit_re_age,
     conf.int = TRUE,
     exponentiate = TRUE,
     effects = "fixed")
```
OR 2.19 (0.94-5.08), p = 0.07.
The OR estimate is higher than that from the GEE analysis. 
rho = ? (from Stata: 0.25)
LRT on rho=0, p = ? (from Stata: 0.001)

Now check the reliability of these estimates with a quadrature check.

I haven't found a function that's the exact equivalent to `quadchk()` in Stata. What I can do is fit two more models with different `nAGQ` values, and checking the difference in the estimates yourself. If the relative differences are < 0.01, the estimates from your model are reasonably reliable.

```{r}
# Original model
logit_re_age %>% fixef()

# Model with 8 quadrature points
glmer(mantoux ~ ix_cough + (1|id),
      data = tb,
      family = "binomial",
      nAGQ = 8) %>% fixef()

# Model with 16 quadrature points
glmer(mantoux ~ ix_cough + (1|id),
      data = tb,
      family = "binomial",
      nAGQ = 16) %>% fixef()
```
All the relative differences are less than 0.01, suggesting that the estimates from the RE model are reasonably reliable.


## 17. Logistic regression, RE, with covariates

Fit a random effects logistic model to estimate the odds of positive Mantoux according to cough in the index case, controlling for:
* HIV status of the index case
* age of the household contact
* household clustering

How would you summarise and interpret your results?

```{r}
logit_re_multi <- glmer(mantoux ~ ix_cough + ix_hiv + agegrp + (1|id),
                        data = tb,
                        family = "binomial",
                        nAGQ = 12)

tidy(logit_re_multi,
     conf.int = TRUE,
     exponentiate = TRUE,
     effects = "fixed")
```
After adjusting for index case HIV, age group, and clustering, the odds of positive Mantoux in contacts of an index case with a cough lasting more than 2 months are 1.88 times higher than in contacts of indec cases with shorter cough duration. (CI 0.80-4.41). There is not good evidence for this association (p = 0.14)
Estimates from Stata are slightly different (OR 1.85, 0.79-4.31, p = 0.15)


## 18. Logistic regression, GEE, with covariates

Do the same as Q17 but with GEE, and compare the results.

```{r}
logit_gee_covar <- geeglm(mantoux ~ ix_cough + ix_hiv + agegrp,
                          data = drop_na(tb, ix_cough),
                          id = id,
                          family = "binomial",
                          corstr = "exchangeable")

# Output
tidy(logit_gee_covar,
     conf.int = TRUE,
     exponentiate = TRUE)
```
 OR 1.67 (0.83, 3.35)
[OR 1.64 (0.81, 3.32) in Stata]


As GEEs are not based on likelihood, we can't use LRTs. Instad, we can use Wald tests of simple and composite linear hypotheses.

*-----*
*ISSUE* I don't know the R equivalent to Stata's `testparm`.
*-----*

What hypotheses are being tested? What do you conclude?

>
```{stata}
testparm 2.agegrp 3.agegrp
```
Test linear hypotheses after estimation
H0: after taking account of... there is no association between age and outcome. 
p >chi2: <0.001

-------------------------------------------------------------------------------
