---
title: "1: Classical analysis of categorical variables"
subtitle: "R 4 ASME"
author: Andrea Mazzella [(github)](https://github.com/andreamazzella)
output: html_notebook
---

-------------------------------------------------------------------------------

## Prerequisites

You'll better understand this code if you know about the following concepts. If you want to learn more about them, you could go through some of my introductory material [here](https://github.com/andreamazzella/IntRo).

* Use of .Rmd notebooks
* Use of packages
* Assignment `<-`
* Pipes `%>%` `%$%` `%<>%`
* Data and variable types
* Operators `=` `|` `&` `~` `$` `[`


## Contents

* Basic data management
  * import
  * explore
  * clean
  * tidy
  * export
* Classical statistical analysis of categorical variables
  * chi-squared test
  * test for trend
  * OR
  * adjusted OR
  * OR with logistic regression

-------------------------------------------------------------------------------

## 0. Load packages

```{r message=FALSE, warning=FALSE}
# Load packages
library("haven")
library("magrittr")
library("epitools")
library("summarytools")
library("pubh")
library("rstatix")
library("tidyverse")

# Limit significant digits to 2, remove scientific notation
options(digits = 2, scipen = 999)
```
NB: {summarytools} can have issues on Macs - you might need to install some other software. If you get an error message, try following the advice that it gives you.

# Basic data management

## 1. Change the default directory

If you use an .Rmd notebook like this one, you don't need to change directory – R assumes the default directory is the directory where the .Rmd is kept.


## 2. Import and explore a .dta dataset

The function `read_dta()` is from package {haven}.
```{r}
mwanza <- read_dta("mwanza.dta")
```

Unfortunately, Stata help files cannot be accessed in R. Key points regarding this dataset:
- Case-control study (not matched)
- Cases: women aged >=15 years living with HIV
- Controls: random sample of HIV-negative women
- Cases and controls were interviewed about potential risk factors for HIV
- All variables are categorical and there are no value labels.

To look at the variables and their types:
```{r}
# Variable names
names(mwanza)

# Variable names, types, and first values
glimpse(mwanza)

# Visualise the whole dataset, including variable labels
View(mwanza)
```


## 3. Function syntax and getting help

The general syntax in R is:
`function(argument1, argument2)`
Arguments are always separated by a comma.
Unlike Stata, R can use multiple datasets at the same time, which means that you always need to specify the dataset:
`function(dataset)`
With package {tidyverse}, this can be written as:
`dataset %>% function()`
```{r}
# These two are equivalent
names(mwanza)
mwanza %>% names()
```

To use a variable, you still need to specify which dataset it is from:
`function(dataset$variable1)`
With package {magrittr}, this can be written as:
`dataset %$% function(variable1)`

For example, to tabulate education by HIV status with column percents, you can use {summarytools}' `ctable()` function:
```{r}
# These two are equivalent
ctable(mwanza$ed, mwanza$case, prop = "c")
mwanza %$% ctable(ed, case, prop = "c")

# Base R
table(mwanza$ed, mwanza$case)
```

Unlike Stata, you can't abbreviate function or variable names. But if you start typing a variable in a dataset, RStudio will guess what it is and you can select it by pressing Tab and then Enter.

You can filter rows and select columns with functions from package {dplyr}, part of the {tidyverse}.
For example, to do the same tabulation but only in those aged less than 30, you would type:
```{r}
mwanza %>%
  filter(age1 <= 3) %$%
  ctable(ed, case, prop = "c")
```

I'm not sure what the equivalent for `by varlist1:` is in R – presumably, you would need some sort of iteration, like a `for` loop, which I think is beyond the scope of ASME.

To get help about a function or a package, you put a question mark and then the name of that function/package:
```{r}
? View
```
There is also a lot of help online - if you get an error, try googling the error.


## 4. Saving the results

Whenever you save the .Rmd notebook, a record will be created in the same folder – its file format will depend on what's written in the "output:" field at the top of the .Rmd; this one produces an html file.


## 5a: Factor variables

R is more efficient if you tell it that variables with values 0/1/2/3 are in fact categorical and not numerical.
```{r}
# Create a vector of categorical variable names
categ <- c("comp", "case", "ed", "eth", "rel", "msta", "bld", "inj", "skin",
           "fsex", "npa", "pa1", "usedc", "ud", "ark", "srk", "ed2")

# Make them all categorical
mwanza[categ] <- lapply(mwanza[categ], as.factor)
```


## 5b. Creating and recoding variables

The variable `age1` is coded in groups:
1: 15-19, 2: 20-24, 3: 25-29, 4: 30-34, 5: 35-44, 6: 45-54

The variable `ed` represents year of education, coded as 1: none, 2: 1-3 years, 3: 4-6 years, 4: 7+ years

The variable `ed2` is binary; 0: none, 1: 1+ years. The variable `npa` represents lifetime number of sexual partners, coded as 1: 0-1, 2: 2-4, 3: 5-9, 4: 10+, 9: missing.

To create a new variable (or change an existing one), you use the function `mutate()`. You can combine this with `as.factor()` and `case_when()` to relevel a categorical variable and give labels to the values in one single bit of code.
You can then crosstabulate with `table(..., useNA = "ifany")` to ensure this went well.
```{r}
# Create a new variable, relevel and label
mwanza %<>%
  mutate(age2 =  as.factor(case_when(age1 <= 2 ~ "15-24",
                                     age1 == 3 | age1 == 4 ~ "25-34",
                                     age1 == 5 | age1 == 6 ~ "35+")))

# Ensure it went ok
mwanza %$% table(age1, age2, useNA = "ifany")
```

Missing values are recorded as NA in R. To tell R to treat the value of 9 as missing for variable `npa`, you have two options:
a) assign the value NA to the subsetted filtered variable; b) use the `na_if()` function from the tidyverse, 
```{r}
# Option a: base R
mwanza$npa[mwanza$npa == 9] <- NA

# Option b: tidyverse
# mwanza %<>% mutate(npa = na_if(npa, 9))

# CHeck it worked ok
mwanza %$% table(npa, useNA = "ifany")
```

To recode more than one variable I assume you need to use iteration, again.
```{r}
# Sorry, this is poor coding!
mwanza %<>% mutate(msta = na_if(msta, 9))
mwanza %$% table(msta, useNA = "ifany")

mwanza %<>% mutate(eth = na_if(eth, 9))
mwanza %$% table(eth, useNA = "ifany")

mwanza %<>% mutate(rel = na_if(rel, 9))
mwanza %$% table(rel, useNA = "ifany")
```


## 6. Saving the current dataset

Don't overwrite the existing dataset, just make a copy.
```{r}
# Stata
mwanza %>% write_dta("mwanza2.dta")
```

-------------------------------------------------------------------------------


# Classical statistical analysis of categorical variables


## 7. Chi-squared testing

Many ways of doing this. `ctable()` can do this by adding an option `chisq = T`. NB: R warns you a chi-square might not be ideal, because there are not many observations per cell. A Fisher test is more appropriate.

What is the effect of education on HIV status?
```{r}
# summarytools
mwanza %$% ctable(ed, case, chisq = T)

# Base R
mwanza %$% chisq.test(ed, case)
mwanza %$% fisher.test(ed, case)
```


## 8. Test for trend

Is there a linear association?
`prop_trend_test()` is from package {rstatix}. Note that occasionally R can give you a very precise output, when Stata would say "p = 0.0000".
```{r}
# Test for trend
mwanza %$% table(case, ed) %>%
  prop_trend_test()
```


## 9. Odds ratios

Unlike Stata, R can calculate OR for each level of exposure with just one command. This is from package {pubh}.
```{r}
# Stratified ORs
odds_trend(case ~ ed, data = mwanza)
```
It looks like more educated people have higher odds of having HIV than less educated people.


## 10. Adjusted odds ratios

Let's now check if this association still remains after accounting for age.
```{r}
# Option A
pubh::mhor(case ~ age2 / ed2, data = mwanza)

# Option B
# mwanza %$% epiDisplay::mhor(case, ed2, age2, graph = F)
```


## 11. Odds ratios using logistic regression

You can also use logistic regression without covariates to get the OR in all level.
`logistic.display()` makes the output of logistic regression easier to read, and more similar to Stata. `broom::tidy()` is an alternative that's more tidyverse-friendly.
```{r}
# Logistic regression
glm(case ~ ed,
    family = "binomial",
    data = mwanza) %>%
  epiDisplay::logistic.display()
```

You can also add a covariate by adding it after `+`.
Note that the output provides not only the adjusted OR, but also the crude OR (without the covariate) 
```{r}
# Logistic regression with covariate
glm(case ~ ed + age2,
    family = "binomial",
    data = mwanza) %>%
  epiDisplay::logistic.display()
```

-------------------------------------------------------------------------------
