---
title: "6: Stratifying by time"
subtitle: "R 4 ASME"
author: Andrea Mazzella [(GitHub)](https://github.com/andreamazzella)
output: html_notebook
---

-------------------------------------------------------------------------------

## What you will learn

* Create a "survival object" to analyse cohort studies
* Calculate incidence rates from a cohort study
* Stratify rates
 - by a categorical variable
 - by a time variable (by doing a Lexis expansion)
 

## Acknowledgements

Thank you Prof Ruth Keogh @LSHTM, for pointing me at the right direction with regards to `survSplit()` and `pyears()`.

-------------------------------------------------------------------------------

## 0. Packages and options

```{r message=FALSE, warning=FALSE}
# Load packages
library("haven")
library("magrittr")
library("survival")
library("tidyverse")

# Limit significant digits to 3, remove scientific notation
options(digits = 3, scipen = 999)
```


## 1. Data management

Import and explore `whitehal.dta`. These are the variables of interest:
* *Outcome*: death from cardiac event (`chd`)
* *Exposure*: job grade (`grade4`, `grade`)
* *Dates*
 * Date of birth: `timebth`
 * Date of entry: `timein`
 * Date of exit: `timeout`
* *Confounder*: smoking status (`smok`)
```{r}
# Import the dataset
whitehall <- read_stata("whitehal.dta")

# Explore data types
glimpse(whitehall)
```

As you can see, there are no value labels, so I'll add them using the dataset help file.
```{r data_management}
# Rename and factorise variables, label values
whitehall %<>%
  mutate(
    id = as.integer(id),
    all = as.integer(all), # NB outcomes must remain numerical
    chd = as.integer(chd),
    grade4 = factor(
      grade4,
      levels = c(1, 2, 3, 4),
      labels = c("admin", "profess", "clerical", "other")
    ),
    smok = factor(
      smok,
      levels = c(1, 2, 3, 4, 5),
      labels = c("never", "ex", "1-14/day", "15-24/day", "25+/day")
    ),
    grade = factor(grade,
                   levels = c(1, 2),
                   labels = c("higher", "lower")),
    cholgrp = factor(cholgrp),
    sbpgrp = factor(sbpgrp))

# Check it worked ok
glimpse(whitehall)

#Summarise
summary(whitehall)
```

-------------------------------------------------------------------------------


## 2. Calculating rates and RR

Calculate rates stratified by exposure (the two grades of employment: `grade`).

You create a survival object with `Surv()`; it contains duration of follow-up and status at end of follow-up. [equivalent to `stset` in Stata]

You then calculate stratified rates `pyears()`: in the formula, first you put the survival object you have just created, and then the stratification; you then pipe this into `summary()` [equivalent to `strate` in Stata].

(NB: `pyears()` automatically scales from days to years - if you don't want this to happen, for example because your Surv object is already set in years, you need to indicate it with argument `scale = 1`.)

```{r}
# Create survival object
surv_white <- whitehall %$% Surv(time = as.numeric(timein) / 365.25, 
                                 time2 = as.numeric(timeout) / 365.25, 
                                 event = chd)
# Calculate rates
pyears(surv_white ~ grade, data = whitehall, scale = 1) %>%
  summary(n = F, rate = T, ci.r = T, scale = 1000)
```


Calculate the cardiac mortality rate ratio in these two job groups:
```{r}
8.8/4.4
```

-------------------------------------------------------------------------------

## 3. Age as timescale

In order to change the timescale to current age, we need to change the `origin` argument in the `Surv()` function to time of birth.
```{r}
# Create survival object
surv_white_age <- whitehall %$% Surv(time = as.numeric(timein) / 365.25, 
                                     time2 = as.numeric(timeout) / 365.25, 
                                     event = chd,
                                     origin = as.numeric(timebth) / 365.25) ###

# Check rates haven't changed
pyears(surv_white_age ~ grade, data = whitehall, scale = 1) %>%
  summary(n = F, rate = T, ci.r = T, scale = 1000)
```

-------------------------------------------------------------------------------


## 4-7. Lexis expansion

Now let's split the follow-up times into intervals that are specific to sidderent agebands.

To check what R is doing, we'll check record 5001 before and after splitting.
```{r}
whitehall %>% filter(id == "5001") %>% select(-2, -(4:7), -(9:11))
```

Use the `survSplit()` function to create 5-year groups of current age between age 50 and 80, and 10-year groups for the youngest and oldest groups [equivalent to `stsplit` in Stata].
```{r}
# Split
white_split <- survSplit(surv_white_age ~ .,
                         data = whitehall,
                         cut = c(40, seq(50, 80, 5), 90),
                         episode = "ageband")
```

What happened to person record 5001?
It has been expanded into five, and two new columns have been added, indicating the ageband.
```{r}
white_split %>% filter(id == "5001") %>% select(-2, -(4:7), -(9:11))
```

The Lexis expansion won't change the original dataset, therefore the information on events and rates isn't corrupted.

```{r}
# Stratify by grade
pyears(surv_white_age ~ grade,
       data = white_split, 
       scale = 1) %>%
  summary(n = F, rate = T, ci.r = T, scale = 1000)
```

-------------------------------------------------------------------------------


## 8. Stratifying by age band

Now we can use this newly created variable to stratify the rates by age band.
What is the effect of age on cardiac-related mortality?
```{r}
# Stratify by ageband
pyears(surv_white_age ~ ageband,
       data = white_split,
       scale = 1) %>%
  summary(n = F, rate = T, ci.r = T)
```

-------------------------------------------------------------------------------


## 9. Further stratification

You can stratify by another categorical variable by adding it after a + in the right hand-side of the formula in `pyears()`. You then change the options in `summary()` to hide events and person-time. There are options to calculate RR instead of showing the risks, but I can't get them to work.


```{r}
# Calculate rates stratified by age and grade
rates_age_grade <- pyears(surv_white_age ~ ageband + grade,
                          data = white_split,
                          scale = 1)

summary(rates_age_grade, n = F, event = F, pyears = F, rate = T, scale = 1000)

# does not work
# summary(rates_age_grade, rr = T, ci.rr = T, n = F, event = F, pyears = F)

# Calculate rate ratios manually
tribble(
  ~age_band, ~RR,
  "2", 2.13 / 0,
  "3", 3.86 / 1.64,
  "4", 3.95 / 2.53,
  "5", 10.17 / 4.67,
  "6", 13.36 / 7.24,
  "7", 8.70 / 14.21,
  "8", 12.72 / 20.44,
  "9", 39.35 / 23.70
  )
```

*Issue* I also don't know how to calculate an overall MH rate ratio, MH χ², and test for interaction [stuff that Stata's `stmh` does as part of the same command]

`epiR::epi.2by2()` apparently can calculate MH rate ratios by setting the option "method" to "cohort.time". The downside: it requires a very specific input, a 3-way table containing cases and person-years stratified by ageband and the other categorical variable, and I'm not sure how to convert the output from `pyears()` into this very specific format without doing it manually.

```{r does not work}
#test <- summary(py_grade_age, n = F)
#epiR::epi.2by2(test)
```

*Workaround*: use regression methods instead

-------------------------------------------------------------------------------


# Optional exercises


## 10. Smoking

Examine the effect of smoking on cardiac-related mortality.
First, relevel the smoking variable into three: never/ex/current smokers.
```{r}
# Check levels
whitehall %$% table(smok, useNA = "ifany")

# Recode
whitehall %<>%
  mutate(smok3 = as.factor(case_when(smok == "never" ~ "never",
                                     smok == "ex" ~ "ex",
                                     smok == "1-14/day" ~ "current",
                                     smok == "15-24/day" ~ "current",
                                     smok == "25+/day" ~ "current")))

# Order levels
whitehall$smok3 <- fct_relevel(whitehall$smok3, "never", "ex", "current")

# Check it worked
whitehall %$% table(smok3, smok)
```

```{r}
# Rates stratified by smoking
pyears(surv_white_age ~ smok3, data = whitehall, scale = 1) %>%
  summary(n = F, rate = T, ci.r = T, scale = 1000)
```
The mortality rate is higher in smokers than in never-smokers.

Does smoke confound the relationship between job grade and cardiac-related mortality?
```{r}
# Create a new survival object
surv_white_age2 <- whitehall %$% Surv(time = as.numeric(timein) / 365.25, 
                                     time2 = as.numeric(timeout) / 365.25, 
                                     event = chd,
                                     origin = as.numeric(timebth) / 365.25)
# Split
white_split <- survSplit(surv_white_age ~ .,
                         data = whitehall,
                         cut = c(40, seq(50, 80, 5), 90),
                         episode = "ageband")

# Stratified rates
pyears(surv_white_age2 ~ smok3 + grade,
       data = white_split2,
       scale = 1) %>% 
  summary(n = F, event = F, pyears = F, rate = T, scale = 1000)

5.1/1.3
7.4/4.5
10.6/6.3
```


-------------------------------------------------------------------------------


## 11. Statifying on three variables

Examine the effect of job grade on cardiac mortality, adjusting for both age and smoking at the same time. What can you conclude?
*ERROR* code breaks, not sure why
```{r}
# Stratified rates
# pyears(surv_white_age2 ~ ageband + grade + smok3,
#       data = white_split2,
#       scale = 1) %>% 
#  summary(n = F, event = F, pyears = F, rate = T, scale = 1000)
```

-------------------------------------------------------------------------------


## 12. Standardised mortality rate

Not sure how to do this
```{r}

```

-------------------------------------------------------------------------------
