---
title: "15: Additive and multiplicative models"
subtitle: "R 4 ASME"
author: Andrea Mazzella [(GitHub)](https://github.com/andreamazzella)
output: html_notebook
---

-------------------------------------------------------------------------------

## Contents

* *Draft.* Need to research additive Poisson.

-------------------------------------------------------------------------------

## Packages and options

```{r libraries}
# Load packages
library("haven")
library("magrittr")
library("broom")
library("mgcv")      # additive models
library("tidyverse")

# Limit significant digits to 3, remove scientific notation
options(digits = 3, scipen = 9)
```

-------------------------------------------------------------------------------

# 1.

(pen and paper question)

# 2.

## 2a. Data import, exploration, and management

Import and explore `goldmine.dta`. It contains data from a historical cohort studies exploring the risk factors for tubercolosis among male gold miners in South Africa.

These are the variables of interest.

**Outcome*:
 - `tb` (tubercolosis diagnosis, 0/1)

**Exposures*:
 - `hiv` (HIV status at entry: 0/1)
 - `silic` (silicosis grad: 0 = "none", 1 = "possible", 2 = "early", 3 = "advanced")
 
**Time*:
 -`entry` (date of entry, formatted as number of days)
 -`exit` (date of exit, formatted as number of days)
 -`dob` (date of birth, formatted as number of days)

```{r import}
# Import the dataset
gold <- read_stata("goldmine.dta") %>% select(tb, hiv, silic, entry, exit)

# Explore data types
glimpse(gold)
```

We need to transform these days into person-years, and to remove observations with a missing silicosis value. Also, let's factorise and label the exposures.

As we won't need to do Lexis expansions or Cox regressions, we don't need to `Surv()` our data.
```{r data management}
gold %<>% mutate(pyears = (exit - entry) / 365.25,
                 hiv = factor(hiv, levels = c(0, 1), labels = c("neg", "pos")),
                 silic = factor(silic,
                                levels = 0:3,
                                labels = c("none", "possible", "early", "advanced"))) %>%
          drop_na(silic) %>% 
          select(-c("entry", "exit"))
```

Let's summarise this subsetted data to calculate cases, person-years and incidence rates of TB according to HIV status and then silicosis.

```{r rates}
# General summary
summary(gold)
sum(gold$pyears)

# Incidence by HIV
gold %>% group_by(hiv) %>% summarise(count = sum(tb),
                                     person_years = sum(pyears),
                                     rate_percent = sum(tb) / sum(pyears) * 100)

# Incidence by silicosis
gold %>% group_by(silic) %>% summarise(count = sum(tb),
                                       person_years = sum(pyears),
                                       rate_percent = sum(tb) / sum(pyears) * 100)
```
Participants were followed up for a maximum of 6 years. Total of 9947 person-years. Incidence rate 1.1% in HIV neg, 4.9% in PLHIV.


## 2b. Crude effect of HIV

Estimate the crude effect of HIV on TB rate with Poisson regression.

```{r}
crude_hiv <- glm(tb ~ hiv + offset(log(pyears)),
                 family = "poisson",
                 gold)

tidy(crude_hiv, exponentiate = T, conf.int = T)
```
HR (HIV+ vs HIV-) 4.59 (3.48-6.10)


## 2c. Crude effect of silicosis

Do the same but for silicosis.

```{r}
crude_silic <- glm(tb ~ silic + offset(log(pyears)),
                   family = "poisson",
                   gold)

tidy(crude_silic, exponentiate = T, conf.int = T)
```
HR (baseline: no silicosis)
- possible silicosis: 1.83 (1.29-2.54)
- early silicosis:    3.53 (2.13-5.53)
- advanced silicosis: 3.58 (2.31-5.35)


## 2d. Biological plausibility

(pen and paper question)


## 2e. Multivariable Poisson with multiplicative effects

Now examine the combined effects of HIV and silicosis in a multivariable model. Let's use first a multiplicative model, and let's assess if there's interaction on the multiplicative scale.

```{r}
# Multiplicative without interaction
multipl <- glm(tb ~ hiv + silic + offset(log(pyears)),
               family = "poisson",
               gold)

# Multiplicative with interaction
mul_int <- glm(tb ~ hiv * silic + offset(log(pyears)),
               family = "poisson",
               gold)

# LRT
epiDisplay::lrtest(multipl, mul_int)

# Output from first model
tidy(multipl, conf.int = T, exponentiate = T)
```
LRT for interaction, p = 0.43. There is no evidence of interaction on the multiplicative scale.
The rate of tubercolosis in HIV-positive men with advacned silicosis was 20 times higher compared to HIV-negative men without silicosis (HR = 4.82 × 4.12 = 19.9)

### 2e2. Comparing expected and observed cases with multiplicative model

Now obtain the predicted number of tubercolosis cases across HIV and silicosis strata from this model.
Note that you can nest `predict()` inside `mutate()`, so you don't have to add a permanent predicted variable to your dataset.

```{r}
gold %>% mutate(predicted_TB = predict(multipl, type = "response")) %>% 
         group_by(hiv, silic) %>%
         summarise(observed_TB = sum(as.numeric(tb)),
                   predicted_TB = sum(predicted_TB),
                   OE_difference = abs(observed_TB - predicted_TB))
```

For all covariates, the observed number of TB cases is similar to the predicted number under the multiplicative model.


## 2f. Multivariable Poisson with additive effects

Now fit the same model, but with additive effects, as there is biological plausibility for additive effects.
Is there evidence of interaction on this scale?

Function `gam()` from package {mgcv} *should* be able to fit (mixed) generalised additive models.
There's also `gam()` from {gam}.

*-----------*
*---ISSUE---* I don't know how to use `gam()` with Poisson and an identity link.
*-----------* Should I remove pyears from the offset?

```{r broken}
library("mgcv")
gold %<>% mutate(tb_rate = (tb / pyears))

?mgcv::gam()

# Additive model without interaction - broken
additive <- mgcv::gam(tb_rate ~ hiv + silic + offset(log(pyears)),
                      family = poisson(link = "identity"),
                      gold)
summary(additive)
coef(additive) %>% exp()

# Additive model with interaction - broken
additive_inter <- mgcv::gam(tb_rate ~ hiv * silic + offset(log(pyears)),
                            family = poisson(link = "identity"),
                            gold)
summary(additive_inter)
coef(additive_inter) %>% exp()

# LRT
epiDisplay::lrtest(additive, additive_inter)

# gam::gam()
library("gam")
additive_2 <- gam::gam(tb ~ hiv + silic,
                     family = poisson(),
                     gold)
?gam::gam()
tidy(additive_2)
```

> Below: Stata stuff
The analysis is performed using the glm command (see Appendix to Session 16) using the rate
(rather than the number of TB cases) as the outcome. First perform a LRT for interaction by
comparing the log likelihood in the models with and without interaction:
```{stata}
gen rate = tb/pyears
glm rate i.hiv i.silic [iweight=pyears], family(poisson) link(identity)
est store A
glm rate i.hiv##i.silic [iweight=pyears], family(poisson) link(identity)
est store B
lrtest A B
```
Stata output for model 1:
------------------------------------------------------------------------------
             |                 OIM
        rate |      Coef.   Std. Err.      z    P>|z|     [95% Conf. Interval]
-------------+----------------------------------------------------------------
       1.hiv |   .0359824   .0043567     8.26   0.000     .0274435    .0445214
             |
       silic |
          1  |   .0090604   .0037984     2.39   0.017     .0016156    .0165052
          2  |   .0272015   .0118625     2.29   0.022     .0039514    .0504516
          3  |   .0383819   .0105738     3.63   0.000     .0176577    .0591062
             |
       _cons |   .0070582   .0011093     6.36   0.000      .004884    .0092324
------------------------------------------------------------------------------
//the link function is the transformation that is done to the outcome before analysis - in the additive model we are not transforming with the link function, but are still using poisson data
The LRT is given by 16.82 [2*(-955.979- -964.390)] which is a chi-square with 3 df, giving
p=0.0008 and so there is strong evidence of interaction on the additive scale. Note that all three
interaction terms are positive and the “main effects” are also positive (so all RRs>1), suggesting
that the effects are more than additive (multiplicative, perhaps).


### 2f2. Comparing expected and observed cases with additive model

We *should* be able to obtain the predicted number of tubercolosis cases, according to HIV status and silicosis, in the additive model, like this.

*-----------*
*---ISSUE---* This relies on a functioning model...
*-----------*

```{r broken2}
# Broken
gold %>% mutate(expected_rate = predict(additive, type = "response")) %>% 
         group_by(hiv, silic) %>%
         summarise(observed_TB = sum(as.numeric(tb)),
                   expected_TB = expected_rate * pyears,
                 # expected_rate = sum(expected_rate),
                   OE_difference = abs(observed_TB - expected_TB))
```


>
```{stata}
preserve
collapse (sum) tb pyears, by(hiv silic)
gen rate=tb/pyears
glm rate i.hiv i.silic [iweight=pyears], family(poisson) link(identity)
predict expectrate
gen tbexpect=expectrate*pyears
list hiv silic tb tbexpect
restore
list hiv silic tb tbexpect
```
[As the model is based on the outcome of a rate the predict command calculates the expected
rate for each of the 8 unique combinations. This predicted rate is then multiplied by the personyears to obtain the expected number of TB cases]
For some covariates, the observed number of TB cases differs markedly from the expected
number under the additive model. In summary, the strong evidence for an interaction (p=0.0008)
and the discrepancies between the observed and expected values for the additive model suggest
that the effects of HIV and silicosis do not combine additively.

-------------------------------------------------------------------------------
